Master's Project Journal

### 2019-04-21 thesis/NetworkImplemenation.tex
start of conc

\section{Chapter Overview}
This chapter will cover the network implemented in the course of the project, beginning with the individual \ac{ADPLL}'s composition. A number of attributes for each design will be used to analyse and compare their performance. The impact of some minor architectural changes will also be covered, before the implementation of test network will be described. This chapter will also highlight some pitfalls encountered in the process. The \ac{HDL} Verilog was selected for use in this project due to my prior exposure to that language. 5 MHz was chosen as the target centre frequency for these \ac{ADPLL}s, during initial testing it was discovered that the waveform output via the Digilent Nexys4 development board's \ac{GPIO} ports began to degrade around the 10 MHz mark, preventing accurate analysis.

@@

The three architectures chosen for implementation represent a progression from entirely an \ac{FPGA} clock driven system to an \ac{ADPLL} with all components asynchronous with respect to that clock. Therefore the first \ac{ADPLL} design features a clocked oscillator and phase detector, specifically the linear frequency design mentioned in Chapter \ref{chap:3} and the clocked phase detector using a state machine and a bi-directional counter. Skipping over Design 2, the final design used was asynchronous in totality, featuring the \acl{RO} and SigNum/\ac{TDL} phase detector. Design 2 sat between these two, using the \ac{RO} as its oscillator, but retained the clocked phase detector seen in Design 1.

@@

\subsection{Generic Components}
As mentioned in the chapter overview, a number of components were used in all \ac{ADPLL}s implemented, as changes in their design would not impact the overall system.
\subsubsection{Error Combiner}
The error combiner in this design had a number of requirements in order to be suitable for use in both a single \ac{ADPLL} as well as a network operating in both uni- and bi-directional modes, alongside modifications which would change the widths of the error signals requiring averaging. The module was implemented on the basis that the \ac{LF} input width would be identical to the phase detector output width.

The error combiner was designed to take in up to four error signals, all of an identical width which defaults to eight, and four weights to apply in the summation process. A weighted summation is performed by multiplying each error signal by the weight and adding the results together. Regardless of the number of input error signals, a division by four is then performed in order to compute the average error, done by shifting the value right by two. In a \ac{HDL}, unlike C programming or similar, a shift requires no computation and is instead implemented by performing a part-selection. As all error signals are two's complemented signed integers, the error combination process is also carried out using signed values. The code implementing the error combiner can be found in \textsc{ErrorCombiner.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:error_combiner}.

@@

\subsubsection{Loop Filter}
Two \acl{LF}s were implemented, both as \ac{IIR} filters. \ac{FIR} was dismissed due to the extra hardware required to implement the calculations within a single clock cycle, as the \ac{LF} is clocked using the oscillator output, and the greater difficulty of gain adjustment compared to an \ac{IIR}. Clocking on using the generated signal ensures that the discrete time integration is only carried out once per phase comparison. A pitfall that may be encountered implementing an \ac{ADPLL} featuring a divider is not clocking the module on the divided clock, which will result in the integration being carried out at the oscillator output frequency.

@@

Figure \ref{fig:integer_lf} contains an \ac{RTL} diagram describing the implementation. Important to note is the shift applied to the result of the multiplication by \acs{kp} which, as the integration is performed with a larger width to avoid accumulator overflow, preserves the intended relationship between proportional and integral gains. The value of $k_p$ and $k_i$ can be computed using the following formulae:
\begin{gather}
\text{Proportional gain} = k_p\times 2^{-path\_width\_change\_p} \\
path\_width\_change\_p = lsh\_width - rsh\_width \\
lsh\_width = ki\_width - kp\_width  \\
rsh\_width = error\_width + kp\_width + lsh\_width - control\_code\_width \\
\text{Integral gain} = k_i\times 2^{-path\_width\_change\_i} \\
path\_width\_change\_i = - rsh\_width
\end{gather}
In the example system shown in Figure \ref{fig:integer_lf} the proportional gain can be computed as $k_p\times 2^{3-8} = \frac{k_p}{32}$ and the integral gain is $k_i\times 2^{-8} = \frac{k_i}{128}$.
The Verilog implementation of the \ac{LF} using integer arithmetic can be found in \textsc{LoopFilter.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:loop_filter}.

@@

Figure \ref{fig:osc2_impl} contains an \ac{RTL} diagram of this oscillator's implementation. As phase error is a two's complement signed value, its addition to the bias must be carried out using signed arithmetic. Provided the bias is greater than the minimum possible phase error, as has been ensured in this case, the result of this addition is positive signed integer. Being a positive quantity, additional logic performing a conversion from a signed to an unsigned representation is required.
\begin{figure}[h]%
    \centering
    \includegraphics[width=0.8\textwidth]{../osc2_impl} 
    \caption[\ac{DCO} \ac{RTL} diagram as implemented]{\ac{DCO} \ac{RTL} diagram as implemented.}
    \label{fig:osc2_impl}
\end{figure}

@@

Figure \ref{fig:updown_ctr} depicts the implementation of this counter. Apart from the aforementioned capping of the measurement range implemented using a pair of comparators and a multiplexer, important to note is the bit width of the increment. Despite overflow being disallowed when the number is interpreted as a signed number, the addition of a 7 bit wide -1 to the accumulator value will cause overflow in the adder. This, however, is perfectly valid as an implementation of subtraction in two's complement. On the right side of the diagram the output register can be seen. Clocked using the \ac{FSM}s measurement interval complete signal, this register preserves the phase error until the termination of the following interval.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../updown}
    \caption[\ac{RTL} diagram of up-down counter]{\ac{RTL} diagram of up-down counter.}
    \label{fig:updown_ctr}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \subfloat[Vivado simulation.]{{\includegraphics[width=0.45\textwidth]{../sim_locking} }}%
    \subfloat[Hardware measurement.]{{\includegraphics[width=0.45\textwidth]{../impl_locking} }}%
    \caption[\ac{ADPLL} 1 Vivado locking simulations]{\ac{ADPLL} 1 Vivado locking simulations.}
    \label{fig:sim_locking}
\end{figure}
In Figure \ref{fig:sim_locking} (a), the startup behaviour of the system can be observed. This simulation was performed with a reference at exactly $5~\si{\mega\hertz}$ and an initial deviation of $252~\si{\kilo\hertz}$, and the plot shows the inverse of the \ac{DCO} period against the time since the oscillator was enabled. After some initial oscillation the period settles around that of the reference. Once locked, cycle-to-cycle jitter was calculated to be $3.491~\si{\nano\second}$ using the standard deviation of the periods, with a worst case cycle-to-cycle change of $9.299~\si{\nano\second}$. Lock was visually determined to have occurred after $20.91~\si{\micro\second}$, after which the oscillator frequency began to change in a repetitive pattern. later the same test carried in hardware with identical initial conditions, the results of which are shown in Figure \ref{fig:sim_locking} (b). This plot is derived from data captured using an Agilent MSO7054A, using a setup that will be further explained in Section \ref{section:measurement_setup}. Despite lacking a non ideal reference, less violent initial oscillation, and a lesser until lock could be visually determined, the simulation provided a good insight into the system's behaviour once implemented on an \ac{FPGA}.

\subsubsection{\acs{ADPLL} 1 Design Summary}
The listed components were brought together in \textsc{NetworkADPLL.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:network_adpll}. Important to note is that each \ac{ADPLL} only contains two \acl{PD}s despite during network operation the need for phase comparisons to be made with up to four neighbours. The two ``missing'' \ac{PD}s are in fact synthesised as part of the other \ac{ADPLL}s, with which the comparison would be made. This allows for a 50\% reduction in the number of \ac{PD}s required, while also ensuring identical error is received by both \ac{ADPLL}s. Each \ac{ADPLL} performs comparison with the \ac{DCO} ``above'' and ``left'' of them in the Cartesian grid, and a port in the module outputs the results for use by the other \ac{ADPLL}s involved. %TODO juvenile wording
Table \ref{table:adpll1} contains a brief overview of the bit widths, tuning ranges and other configuration information for this \ac{ADPLL}.
\begin{table}[!h]
    \begin{center}
        \begin{tabular}{|l|r|r|r|}
            \multicolumn{4}{c}{\ac{DCO} Tuning Range} \T\\
            \hline
            \multicolumn{1}{|c|}{-} & \multicolumn{1}{c|}{Minimum} & \multicolumn{1}{c|}{Step} & \multicolumn{1}{c|}{Maximum} \T\\
            \hline
            Frequency & $1.008~\si{\mega\hertz}$ & \multicolumn{1}{r|}{$62.988~\si{\kilo\hertz}$} & $8.944~\si{\mega\hertz}$ \T\\
            \hline
            Period & $992.0~\si{\nano\second}$ & \multicolumn{1}{r|}{$2.5~\si{\nano\second}$ (at $5~\si{\mega\hertz}$)} & $111.8~\si{\nano\second}$ \T\\
            \hline
        \end{tabular}
        \begin{tabular}{|l|r|l|r|}
            \multicolumn{4}{c}{Configuration as Implemented} \T\\
            \hline
            \ac{DCO} Counter Width & 12 bits & \ac{DCO} Control Width (max) & 12 bits \T\\
            \hline
            \ac{DCO} Bias Point & 79 & Runtime Gains & Enabled \T\\
            \hline
            Phase Detector Error Width & 7 & Error Sum Weight Width & 4 \T\\
            \hline
            \acs{kp} width & 5 & \acs{ki} & 8 \T\\
            \hline
            \acs{kp} & $2^{-5}\times[0,15]\cap\mathbb{Z}$ & \acs{ki} & $2^{-8}\times[0,15]\cap\mathbb{Z}$ \T\\
            \hline
            Detection Resolution & $3.875~\si{\nano\second}$ & Detection Phase Resolution & $6.975\si{\degree}$ (at $5~\si{\mega\hertz}$)\\
            \hline
        \end{tabular}
    \end{center}
\caption[ADPLL Design 1 Summary]{ADPLL Design 1 Summary.}
\label{table:adpll1}
\end{table}

\subsection{\acs{ADPLL} Design 2}
The second \ac{ADPLL} design implemented during the course of this project was somewhat of a stepping stone between fully synchronous and asynchronous to the \ac{FPGA} clock. The clocked phase detector is reused from above, albeit with an altered error width, while the \ac{DCO} is replaced by a \acl{RO}. While the phase detector was known to work due to testing in \ac{ADPLL} 1, the \ac{DCO} could not be simulated using accurate timings so all testing had to be done in hardware. This presented a challenge, as depending on the implementation, the characteristics of the \ac{DCO} could change, most importantly the centre frequency. The module implementing this \ac{ADPLL} can be found in \textsc{NetworkRingADPLL.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:network_ring_adpll}.

\subsubsection{\acl{DCO}}
\ac{ADPLL} 2 exchanges the \ac{FPGA} clocked oscillator for \ac{RO}, with the aim of obtaining performance more akin to that of an \ac{ASIC} based mixed-signal implementation, due to the variation in period steps both between oscillators and within the same oscillator due to the oscillator layout. Also advantageous is the improvement in period resolution obtained using this design, when compared to that of the \ac{FPGA} clocked design. Taking the formula given in Chapter \ref{chap:3} and the inverter propagation time used by Vivado in post synthesis/implementation simulations of $315~\si{\pico\second}$, the period step can be computed:
\begin{align}
t_{step} &= \text{inverters per step}\times\text{inversions per period}\times\text{propagation delay} \\
t_{step} &= \text{inverters per step}\times\text{inversions per period}\times\tau_{inv} \\
t_{step} &= 2\times 2\times 315~\si{\pico\second} = 1.26~\si{\nano\second}
\end{align}
Using the propagation delay, $\tau_{inv}$, the number of inverters required to produce the signal of period $200~\si{\nano\second}$ can easily be computed:
\begin{equation}
\text{num\_inverters} = \left \lfloor{ T_{osc~@~5~\si{\mega\hertz}}\times \frac{1}{2\times\tau_{inv}}}\right \rceil = \left \lfloor{ \frac{200~\si{\nano\second}}{0.63~\si{\nano\second}}}\right \rceil = 317
\end{equation}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../ro_new}
    \caption[\acl{RO} RTL diagram]{\acl{RO} RTL diagram.}
    \label{fig:ro_impl}
\end{figure}
Figure \ref{fig:ro_impl} contains an \ac{RTL} diagram of the \ac{RO} as implemented. The inverter chain is specified with a maximum length of $n$ and a maximum number of removable inverters of $k$. In order to preserve the unstable circuit, by maintaining an odd number of inverters, only multiples of two inverters may be removed at a time. The already introduced \ac{DCO} and \ac{PFD} designs all followed the same convention regarding phase error and the corresponding impact on the control code: In the \ac{PFD}, if the generated signal was lagging the reference, meaning it should ``go faster'' in order to ``catch up'', the error given a positive sign. Similarly in the \ac{FPGA} clocked \ac{DCO}, a positive change in the magnitude of the control code resulted in a higher frequency. In the interests of consistency and extensibility the same behaviour was implemented for this \ac{DCO} also, with the number of inverters in the chain given as:
\begin{equation}
\text{num inverters} = n - 2\times CC,~CC<\frac{k}{2}
\end{equation}
Where $CC$ is the unsigned integer control code input. $k$ is determined by the bit width of the control code which, as in the case of the clocked \ac{DCO}, is also an unsigned number. An inverter chain will oscillate freely and cannot be disabled. To this end the first inverter in the chain was replaced by a \acs{NAND} gate to grant control over its operation.

@@

As mentioned in previous chapters, the absence of direct control over layout can lead to significant variation of the propagation time though inverters. This may occur in three ways: Firstly, the layout of individual \ac{RO}s may be significantly different, thus resulting in poor overlap of tuning regions. Anecdotally, while implementing a 3x3 network, 7 of the 9 oscillators had centre frequencies within a $100~\si{\kilo\hertz}$ span but two lay more than $500~\si{\kilo\hertz}$ away in opposite directions which prevented the network from locking. Secondly, within an \ac{RO} the propagation delay between each inverter may vary, which results in a variable frequency step, possibly changing the locking range. These two effects represent an extreme version of the variation due to process or manufacturing seen on \ac{ASIC}s. The final variation is possibly the most frustrating, and occurs when between implementations the \ac{EDA}, Vivado in the case of this project, changes the layout of the \ac{RO}. This may occur as a result of a direct change, or as a result of seemingly innocuous changes to unrelated modules. While the final problem can be avoided, once the performance of the \ac{RO} is satisfactory, by locking down module, the remaining two problems can only be mitigated somewhat.

@@

When, later in the project, additional \ac{RO}s were implemented to form networks, 373 inverters proved a suitable starting point which consistently delivered \ac{DCO}s with a centre frequency within $200~\si{\kilo\hertz}$ of $5~\si{\mega\hertz}$. For the purposes of describing a typical tuning range, the average propagation delay of an \ac{RO} consisting of 373 inverters with a centre frequency of $5.06~\si{\mega\hertz}$ will be used. To avoid the aforementioned issues with constant control codes, the fixed code was achieved by implementing an \ac{ADPLL} and setting the \ac{PI} filter gains to zero at runtime before performing a reset.

@@

\subsection{\acs{ADPLL} Design 3}
\acs{ADPLL} Design 3 has no dependence on the clock provided by the \ac{FPGA}'s clock management utility, as both the \ac{RO} and the \ac{TDC} are designed using inverters, while a Bang-Bang detector replaces the \ac{FPGA} clock reliant \ac{FSM} of the two previous designs. This introduces a second source of implementation based variation between \ac{ADPLL}s, however, this has significantly lesser impact on the frequency range of the oscillator and no impact on the centre frequencies. The lack of an \ac{FPGA} clocked element results means this design will have the closest behaviour to that of an \ac{ASIC} based \ac{ADPLL} implementation, although as the level of variation due to implementation is much larger on an \ac{FPGA}, the impact of non-idealities will be exaggerated.
%same interface
In order to allow quick exchange of \ac{PD}s or other modules, the module in which this \ac{ADPLL} is implemented is still retains the \ac{FPGA} clock as an input. In fact, this module shares its implementation with that of \ac{ADPLL} 2, as while the \ac{PD} designs are drastically different, the interface is identical. The module implementing this \ac{ADPLL} can be found in \textsc{NetworkRingADPLL.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:network_ring_adpll}.

@@

\subsubsection{Phase Detector}
Instead of an \ac{FPGA} clocked phase detector, \ac{ADPLL} 3 uses the modified Bang-Bang detector and \ac{TDL} approach to phase comparison seen in Section \ref{section:signum}. Sign detection is implemented exactly as described in that section, with the double \ac{SR} latch approach used to determine the sign of the phase error. A flip-flop is inserted after the second \ac{SR} latch, clocked on the signal \texttt{done}, which holds the output value and ensures that the sign changes at the same time as the magnitude. The previous discussion of this phase detector proposed no implementation of the edge detectors, which are easily implemented using flip-flops clocked on the signal of interest, with a constant $1$ on the $D$ input. In this form the edge detector is single use only, as subsequent edges will not cause any change in the flip-flop output. This is important as it ensures the measurement interval will only terminate when a rising edge is seen on the signal that did not start the measurement interval, thus avoiding the mode locking seen in the original implementation of the \ac{FPGA} clocked \ac{PD}. Once the measurement interval has completed, the \texttt{done} signal is used to reset the edge detectors asynchronously. It is possible that an edge may occur within the time taken to reset the counters, and thus the detector incorrectly gauge the start of a measurement interval, however this can only occur in exceptional circumstances in the acquisition period. Therefore it is possible to ignore this scenario, as the only impact will be an increased time taken to acquire a lock.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../new_pdet1}
    \caption[Modified Bang-Bang sign detector RTL diagrams]{Modified Bang-Bang sign detector RTL diagrams.}
    \label{fig:signdet_impl}
\end{figure}

>>>>

\section{Chapter Overview}
This chapter will cover the network implemented in the course of the project, beginning with the individual \ac{ADPLL}'s composition. A number of attributes for each design will be used to analyse and compare their performance. The impact of some minor architectural changes will also be covered, before the implementation of test network will be described. This chapter will also highlight some pitfalls encountered in the process. The \ac{HDL} Verilog was selected for use in this project due to my prior exposure to that language. 5 MHz was chosen as the target centre frequency for these \acp{ADPLL}, during initial testing it was discovered that the waveform output via the Digilent Nexys4 development board's \ac{GPIO} ports began to degrade around the 10 MHz mark, preventing accurate analysis.

@@

The three architectures chosen for implementation represent a progression from entirely \iac{FPGA} clock driven system to \iac{ADPLL} with all components asynchronous with respect to that clock. Therefore the first \ac{ADPLL} design features a clocked oscillator and phase detector, specifically the linear frequency design mentioned in Chapter \ref{chap:3} and the clocked phase detector using a state machine and a bi-directional counter. Skipping over Design 2, the final design used was asynchronous in totality, featuring the \acl{RO} and SigNum/\ac{TDL} phase detector. Design 2 sat between these two, using the \ac{RO} as its oscillator, but retained the clocked phase detector seen in Design 1.

@@

\subsection{Generic Components}
As mentioned in the chapter overview, a number of components were used in all \acp{ADPLL} implemented, as changes in their design would not impact the overall system.
\subsubsection{Error Combiner}
The error combiner in this design had a number of requirements in order to be suitable for use in both a single \ac{ADPLL} as well as a network operating in both uni- and bi-directional modes, alongside modifications which would change the widths of the error signals requiring averaging. The module was implemented on the basis that the \ac{LF} input width would be identical to the phase detector output width.

The error combiner was designed to take in up to four error signals, all of an identical width which defaults to eight, and four weights to apply in the summation process. A weighted summation is performed by multiplying each error signal by the weight and adding the results together. Regardless of the number of input error signals, a division by four is then performed in order to compute the average error, done by shifting the value right by two. In \iac{HDL}, unlike C programming or similar, a shift requires no computation and is instead implemented by performing a part-selection. As all error signals are two's complemented signed integers, the error combination process is also carried out using signed values. The code implementing the error combiner can be found in \textsc{ErrorCombiner.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:error_combiner}.

@@

\subsubsection{Loop Filter}
Two \aclp{LF} were implemented, both as \ac{IIR} filters. \ac{FIR} was dismissed due to the extra hardware required to implement the calculations within a single clock cycle, as the \ac{LF} is clocked using the oscillator output, and the greater difficulty of gain adjustment compared to \iac{IIR}. Clocking on using the generated signal ensures that the discrete time integration is only carried out once per phase comparison. A pitfall that may be encountered implementing \iac{ADPLL} featuring a divider is not clocking the module on the divided clock, which will result in the integration being carried out at the oscillator output frequency.

@@

Figure \ref{fig:integer_lf} contains \iac{RTL} diagram describing the implementation. Important to note is the shift applied to the result of the multiplication by \acs{kp} which, as the integration is performed with a larger width to avoid accumulator overflow, preserves the intended relationship between proportional and integral gains. The value of $k_p$ and $k_i$ can be computed using the following formulae:
\begin{gather}
\text{Proportional gain} = k_p\times 2^{-path\_width\_change\_p} \\
path\_width\_change\_p = lsh\_width - rsh\_width \\
lsh\_width = ki\_width - kp\_width  \\
rsh\_width = error\_width + kp\_width + lsh\_width - control\_code\_width \\
\text{Integral gain} = k_i\times 2^{-path\_width\_change\_i} \\
path\_width\_change\_i = - rsh\_width
\end{gather}
In the example system shown in Figure \ref{fig:integer_lf} the proportional gain can be computed as $k_p\times 2^{3-8} = \frac{k_p}{32}$ and the integral gain is $k_i\times 2^{-8} = \frac{k_i}{128}$.
The Verilog implementation of the \ac{LF} using integer arithmetic can be found in \textsc{LoopFilter.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:loop_filter}.

@@

Figure \ref{fig:osc2_impl} contains \iac{RTL} diagram of this oscillator's implementation. As phase error is a two's complement signed value, its addition to the bias must be carried out using signed arithmetic. Provided the bias is greater than the minimum possible phase error, as has been ensured in this case, the result of this addition is positive signed integer. Being a positive quantity, additional logic performing a conversion from a signed to an unsigned representation is required.
\begin{figure}[h]%
    \centering
    \includegraphics[width=0.8\textwidth]{../osc2_impl} 
    \caption[\ac{DCO} \ac{RTL} diagram as implemented]{\ac{DCO} \ac{RTL} diagram as implemented.}
    \label{fig:osc2_impl}
\end{figure}

@@

Figure \ref{fig:updown_ctr} depicts the implementation of this counter. Apart from the aforementioned capping of the measurement range implemented using a pair of comparators and a multiplexer, important to note is the bit width of the increment. Despite overflow being disallowed when the number is interpreted as a signed number, the addition of a 7 bit wide -1 to the accumulator value will cause overflow in the adder. This, however, is perfectly valid as an implementation of subtraction in two's complement. On the right side of the diagram the output register can be seen. Clocked using the \acp{FSM} measurement interval complete signal, this register preserves the phase error until the termination of the following interval.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../updown}
    \caption[\ac{RTL} diagram of up-down counter]{\ac{RTL} diagram of up-down counter.}
    \label{fig:updown_ctr}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \subfloat[Vivado simulation.]{{\includegraphics[width=0.45\textwidth]{../sim_locking} }}%
    \subfloat[Hardware measurement.]{{\includegraphics[width=0.45\textwidth]{../impl_locking} }}%
    \caption[\ac{ADPLL} 1 Vivado locking simulations]{\ac{ADPLL} 1 Vivado locking simulations.}
    \label{fig:sim_locking}
\end{figure}
In Figure \ref{fig:sim_locking} (a), the startup behaviour of the system can be observed. This simulation was performed with a reference at exactly $5~\si{\mega\hertz}$ and an initial deviation of $252~\si{\kilo\hertz}$, and the plot shows the inverse of the \ac{DCO} period against the time since the oscillator was enabled. After some initial oscillation the period settles around that of the reference. Once locked, cycle-to-cycle jitter was calculated to be $3.491~\si{\nano\second}$ using the standard deviation of the periods, with a worst case cycle-to-cycle change of $9.299~\si{\nano\second}$. Lock was visually determined to have occurred after $20.91~\si{\micro\second}$, after which the oscillator frequency began to change in a repetitive pattern. later the same test carried in hardware with identical initial conditions, the results of which are shown in Figure \ref{fig:sim_locking} (b). This plot is derived from data captured using an Agilent MSO7054A, using a setup that will be further explained in Section \ref{section:measurement_setup}. Despite lacking a non ideal reference, less violent initial oscillation, and a lesser until lock could be visually determined, the simulation provided a good insight into the system's behaviour once implemented on \iac{FPGA}.

\subsubsection{\acs{ADPLL} 1 Design Summary}
The listed components were brought together in \textsc{NetworkADPLL.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:network_adpll}. Important to note is that each \ac{ADPLL} only contains two \aclp{PD} despite during network operation the need for phase comparisons to be made with up to four neighbours. The two ``missing'' \acp{PD} are in fact synthesised as part of the other \acp{ADPLL}, with which the comparison would be made. This allows for a 50\% reduction in the number of \acp{PD} required, while also ensuring identical error is received by both \acp{ADPLL}. Each \ac{ADPLL} performs comparison with the \ac{DCO} ``above'' and ``left'' of them in the Cartesian grid, and a port in the module outputs the results for use by the other \acp{ADPLL} involved. %TODO juvenile wording
Table \ref{table:adpll1} contains a brief overview of the bit widths, tuning ranges and other configuration information for this \ac{ADPLL}.
\begin{table}[!h]
    \begin{center}
        \begin{tabular}{|l|r|r|r|}
            \multicolumn{4}{c}{\ac{DCO} Tuning Range} \T\\
            \hline
            \multicolumn{1}{|c|}{-} & \multicolumn{1}{c|}{Minimum} & \multicolumn{1}{c|}{Step} & \multicolumn{1}{c|}{Maximum} \T\\
            \hline
            Frequency & $1.008~\si{\mega\hertz}$ & \multicolumn{1}{r|}{$62.988~\si{\kilo\hertz}$} & $8.944~\si{\mega\hertz}$ \T\\
            \hline
            Period & $992.0~\si{\nano\second}$ & \multicolumn{1}{r|}{$2.5~\si{\nano\second}$ (at $5~\si{\mega\hertz}$)} & $111.8~\si{\nano\second}$ \T\\
            \hline
        \end{tabular}
        \begin{tabular}{|l|r|l|r|}
            \multicolumn{4}{c}{Configuration as Implemented} \T\\
            \hline
            \ac{DCO} Counter Width & 12 bits & \ac{DCO} Control Width (max) & 12 bits \T\\
            \hline
            \ac{DCO} Bias Point & 79 & Runtime Gains & Enabled \T\\
            \hline
            Phase Detector Error Width & 7 & Error Sum Weight Width & 4 \T\\
            \hline
            \acs{kp} width & 5 & \acs{ki} & 8 \T\\
            \hline
            \acs{kp} & $2^{-5}\times[0,15]\cap\mathbb{Z}$ & \acs{ki} & $2^{-8}\times[0,15]\cap\mathbb{Z}$ \T\\
            \hline
            Detection Resolution & $3.875~\si{\nano\second}$ & Detection Phase Resolution & $6.975\si{\degree}$ (at $5~\si{\mega\hertz}$)\\
            \hline
        \end{tabular}
    \end{center}
\caption[ADPLL Design 1 Summary]{ADPLL Design 1 Summary.}
\label{table:adpll1}
\end{table}

\subsection{\acs{ADPLL} Design 2}
The second \ac{ADPLL} design implemented during the course of this project was somewhat of a stepping stone between fully synchronous and asynchronous to the \ac{FPGA} clock. The clocked phase detector is reused from above, albeit with an altered error width, while the \ac{DCO} is replaced by \iacl{RO}. While the phase detector was known to work due to testing in \ac{ADPLL} 1, the \ac{DCO} could not be simulated using accurate timings so all testing had to be done in hardware. This presented a challenge, as depending on the implementation, the characteristics of the \ac{DCO} could change, most importantly the centre frequency. The module implementing this \ac{ADPLL} can be found in \textsc{NetworkRingADPLL.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:network_ring_adpll}.

\subsubsection{\acl{DCO}}
\ac{ADPLL} 2 exchanges the \ac{FPGA} clocked oscillator for \ac{RO}, with the aim of obtaining performance more akin to that of \iac{ASIC} based mixed-signal implementation, due to the variation in period steps both between oscillators and within the same oscillator due to the oscillator layout. Also advantageous is the improvement in period resolution obtained using this design, when compared to that of the \ac{FPGA} clocked design. Taking the formula given in Chapter \ref{chap:3} and the inverter propagation time used by Vivado in post synthesis/implementation simulations of $315~\si{\pico\second}$, the period step can be computed:
\begin{align}
t_{step} &= \text{inverters per step}\times\text{inversions per period}\times\text{propagation delay} \\
t_{step} &= \text{inverters per step}\times\text{inversions per period}\times\tau_{inv} \\
t_{step} &= 2\times 2\times 315~\si{\pico\second} = 1.26~\si{\nano\second}
\end{align}
Using the propagation delay, $\tau_{inv}$, the number of inverters required to produce the signal of period $200~\si{\nano\second}$ can easily be computed:
\begin{equation}
\text{num\_inverters} = \left \lfloor{ T_{osc~@~5~\si{\mega\hertz}}\times \frac{1}{2\times\tau_{inv}}}\right \rceil = \left \lfloor{ \frac{200~\si{\nano\second}}{0.63~\si{\nano\second}}}\right \rceil = 317
\end{equation}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../ro_new}
    \caption[\acl{RO} RTL diagram]{\acl{RO} RTL diagram.}
    \label{fig:ro_impl}
\end{figure}
Figure \ref{fig:ro_impl} contains \iac{RTL} diagram of the \ac{RO} as implemented. The inverter chain is specified with a maximum length of $n$ and a maximum number of removable inverters of $k$. In order to preserve the unstable circuit, by maintaining an odd number of inverters, only multiples of two inverters may be removed at a time. The already introduced \ac{DCO} and \ac{PFD} designs all followed the same convention regarding phase error and the corresponding impact on the control code: In the \ac{PFD}, if the generated signal was lagging the reference, meaning it should ``go faster'' in order to ``catch up'', the error given a positive sign. Similarly in the \ac{FPGA} clocked \ac{DCO}, a positive change in the magnitude of the control code resulted in a higher frequency. In the interests of consistency and extensibility the same behaviour was implemented for this \ac{DCO} also, with the number of inverters in the chain given as:
\begin{equation}
\text{num inverters} = n - 2\times CC,~CC<\frac{k}{2}
\end{equation}
Where $CC$ is the unsigned integer control code input. $k$ is determined by the bit width of the control code which, as in the case of the clocked \ac{DCO}, is also an unsigned number. An inverter chain will oscillate freely and cannot be disabled. To this end the first inverter in the chain was replaced by \iacs{NAND} gate to grant control over its operation.

@@

As mentioned in previous chapters, the absence of direct control over layout can lead to significant variation of the propagation time though inverters. This may occur in three ways: Firstly, the layout of individual \acp{RO} may be significantly different, thus resulting in poor overlap of tuning regions. Anecdotally, while implementing a 3x3 network, 7 of the 9 oscillators had centre frequencies within a $100~\si{\kilo\hertz}$ span but two lay more than $500~\si{\kilo\hertz}$ away in opposite directions which prevented the network from locking. Secondly, within \iac{RO} the propagation delay between each inverter may vary, which results in a variable frequency step, possibly changing the locking range. These two effects represent an extreme version of the variation due to process or manufacturing seen on \acp{ASIC}. The final variation is possibly the most frustrating, and occurs when between implementations the \ac{EDA}, Vivado in the case of this project, changes the layout of the \ac{RO}. This may occur as a result of a direct change, or as a result of seemingly innocuous changes to unrelated modules. While the final problem can be avoided, once the performance of the \ac{RO} is satisfactory, by locking down module, the remaining two problems can only be mitigated somewhat.

@@

When, later in the project, additional \acp{RO} were implemented to form networks, 373 inverters proved a suitable starting point which consistently delivered \acp{DCO} with a centre frequency within $200~\si{\kilo\hertz}$ of $5~\si{\mega\hertz}$. For the purposes of describing a typical tuning range, the average propagation delay of \iac{RO} consisting of 373 inverters with a centre frequency of $5.06~\si{\mega\hertz}$ will be used. To avoid the aforementioned issues with constant control codes, the fixed code was achieved by implementing \iac{ADPLL} and setting the \ac{PI} filter gains to zero at runtime before performing a reset.

@@

\subsection{\acs{ADPLL} Design 3}
\acs{ADPLL} Design 3 has no dependence on the clock provided by the \ac{FPGA}'s clock management utility, as both the \ac{RO} and the \ac{TDC} are designed using inverters, while a Bang-Bang detector replaces the \ac{FPGA} clock reliant \ac{FSM} of the two previous designs. This introduces a second source of implementation based variation between \acp{ADPLL}, however, this has significantly lesser impact on the frequency range of the oscillator and no impact on the centre frequencies. The lack of \iac{FPGA} clocked element results means this design will have the closest behaviour to that of \iac{ASIC} based \ac{ADPLL} implementation, although as the level of variation due to implementation is much larger on \iac{FPGA}, the impact of non-idealities will be exaggerated.
%same interface
In order to allow quick exchange of \acp{PD} or other modules, the module in which this \ac{ADPLL} is implemented is still retains the \ac{FPGA} clock as an input. In fact, this module shares its implementation with that of \ac{ADPLL} 2, as while the \ac{PD} designs are drastically different, the interface is identical. The module implementing this \ac{ADPLL} can be found in \textsc{NetworkRingADPLL.v}, attached in Appendix \ref{adx:code} Listing \ref{lst:network_ring_adpll}.

@@

\subsubsection{Phase Detector}
Instead of \iac{FPGA} clocked phase detector, \ac{ADPLL} 3 uses the modified Bang-Bang detector and \ac{TDL} approach to phase comparison seen in Section \ref{section:signum}. Sign detection is implemented exactly as described in that section, with the double \ac{SR} latch approach used to determine the sign of the phase error. A flip-flop is inserted after the second \ac{SR} latch, clocked on the signal \texttt{done}, which holds the output value and ensures that the sign changes at the same time as the magnitude. The previous discussion of this phase detector proposed no implementation of the edge detectors, which are easily implemented using flip-flops clocked on the signal of interest, with a constant $1$ on the $D$ input. In this form the edge detector is single use only, as subsequent edges will not cause any change in the flip-flop output. This is important as it ensures the measurement interval will only terminate when a rising edge is seen on the signal that did not start the measurement interval, thus avoiding the mode locking seen in the original implementation of the \ac{FPGA} clocked \ac{PD}. Once the measurement interval has completed, the \texttt{done} signal is used to reset the edge detectors asynchronously. It is possible that an edge may occur within the time taken to reset the counters, and thus the detector incorrectly gauge the start of a measurement interval, however this can only occur in exceptional circumstances in the acquisition period. Therefore it is possible to ignore this scenario, as the only impact will be an increased time taken to acquire a lock.
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../new_pdet1}
    \caption[Modified Bang-Bang sign detector RTL diagrams]{Modified Bang-Bang sign detector RTL diagrams.}
    \label{fig:signdet_impl}
\end{figure}
